## Template for README.Rmd file for paired end RNA-Seq with clumpify step to remove optical duplicates on NovaSeq.  Run add_readme.R to replace @VARIABLES

---
title: "STAR alignments for @RUN"
date: '`r gsub("  ", " ", format(Sys.time(), "%B %e, %Y"))`'
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, comment="", collapse=TRUE)
```

This guide describes the steps used to run [STAR] and other bioinformatic tools
on the Fastq files in GNomEx. All the commands listed in steps #2 to #7 are
organized in a `cmd.txt` file,  which is used to send jobs to the [CHPC]
clusters following the [Pysano] reference manual.  This file also includes
[shell] scripts to parse sample names from FASTQ files and then read and write
to different output files with that prefix.

### 1. Create reference

Download the @NAME FASTA and GTF file from [Ensembl] release @VERSION and
run [STAR] with the `genomeGenerate` option to create the reference database.
The `sjdbGTFfile` option extracts splice junctions from the GTF file with a
maximum possible overhang of @SJDB bases (for @LENGTH bp reads).

```{bash star_ref, eval=FALSE}
wget ftp://ftp.ensembl.org/pub/release-@VERSION/fasta/@LOWER_SPECIES/dna/@SPECIES.@ASSEMBLY.dna.@LEVEL.fa.gz
wget ftp://ftp.ensembl.org/pub/release-@VERSION/gtf/@LOWER_SPECIES/@SPECIES.@ASSEMBLY.@VERSION.gtf.gz
gunzip *.gz

STAR --runMode genomeGenerate \
     --genomeDir star@LENGTH \
     --runThreadN @NCPU \
     --genomeFastaFiles @SPECIES.@ASSEMBLY.dna.@LEVEL.fa \
     --sjdbGTFfile @SPECIES.@ASSEMBLY.@VERSION.gtf \
     --sjdbOverhang @SJDB
```

### 2. Remove optical duplicates

Optical duplicates are a common [problem] on NovaSeq runs and can be removed
before aligning using [clumpify].  Run the tool with the default settings using
`BBmap` version v38.34 (see `optical_dups.txt` for total reads removed).  Note that
samples on the NovaSeq are often split into two different lanes and the code in
`cmd.txt` will first join all reads into a single `@SAMPLE.cat_R1.fq.gz` or `R2` file below.


```{bash clump, eval=FALSE}
clumpify.sh in1=@SAMPLE.cat_R1.fq.gz in2=@SAMPLE.cat_R2.fq.gz \
  out1=@SAMPLE.clump1.fq.gz out2=@SAMPLE.clump2.fq.gz \
  dupedist=12000 dedupe=t optical=t
```

### 3. Trim adapters

Trim the Illumina adpater sequence using [cutadapt] version 1.16.  The -O
option starts trimming after 6 matching bases and -m option will discard trimmed
reads shorter than 20 bases

```{bash cutadapt, eval=FALSE}
cutadapt -j @NCPU -O 6 -m 20 \
  -a AGATCGGAAGAGCACACGTCTGAACTCCAGTCA \
  -A AGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGT \
  -o @SAMPLE.1.fq -p @SAMPLE.2.fq \
  @SAMPLE.clump1.fq.gz @SAMPLE.clump2.fq.gz
```

### 4. Run FastQC

Run [FastQC] on the trimmed fastq file.  This ouputs sequence length distributions,
overrepresented sequences and many other statistics.

```{bash fastq, eval=FALSE}
fastqc -f fastq @SAMPLE.1.fq
fastqc -f fastq @SAMPLE.2.fq
```

### 5. Align reads

Align the trimmed reads using [STAR] version version @STAR to the @NAME
reference in two pass mode and output a BAM file sorted by
coordinates and unstranded bedGraph file.
The `quantMode` option outputs alignments to transcript coordinates for [RSEM].

```{bash star_run, eval=FALSE}
STAR --genomeDir star@LENGTH \
     --runThreadN @NCPU \
     --readFilesIn @SAMPLE.1.fq @SAMPLE.2.fq \
     --twopassMode Basic \
     --outSAMtype BAM SortedByCoordinate \
     --quantMode TranscriptomeSAM \
     --outWigType bedGraph \
     --outWigStrand Unstranded
```

### 6. Count features

Rename the STAR output BAM file and run [featureCounts] version 1.6.3 to count uniquely
aligned reads overlapping features in the GTF file.

```{bash counts, eval=FALSE}
mv Aligned.sortedByCoord.out.bam @SAMPLE.bam
featureCounts -T @NCPU -p -s 2  --largestOverlap -a @SPECIES.@ASSEMBLY.@VERSION.gtf @SAMPLE.counts @SAMPLE.bam
```

STAR reports unique reads and reads that map to 2 to 10 locations on the genome.  These multi-mapping
reads can be assigned to features using [RSEM], which returns the expected counts to transcripts and genes
in `@SAMPLE.isoforms.results` or `@SAMPLE.genes.results`.  These count tables are not used
in the differential expression analyses unless requested.


### 7. Check quality

Run Picard [CollectRnaSeqMetrics] to count the number of reads matching exons,
UTRs, introns and intergenic regions and to calculate the normalized gene
coverage from the top 1000 expressed transcripts. Finally, run [samtools idxstats] to
count the number of reads per chromosome.

### 8. Summarize alignments

[MultiQC] searches the @ALIGNDIR directory for analysis logs and compiles an
HTML report that includes interactive summary tables and plots for all commands
in the `cmd.txt` file.

```{r multiqc, eval=FALSE}
multiqc @ALIGNDIR
```

The General Statistics table at the top of the report includes summaries from
FastQC like total sequences (M Seqs), STAR (M Aligned), featureCounts (M
Assigned) and collectRNASeqMetrics (% rRNA and mRNA).   The remaining sections
summarize outputs from each program.

### 9.  View alignments

Load a BAM file into a database browser like [IGV] by clicking the URL link icon
next to the BAM file name in GNomEx.   Copy and paste the link into IGV by
selecting "File, Load from URL".   Search for a gene or zoom into a specific
region to check alignments. If reads do not align with annotations, make sure
the correct reference assembly is selected (hg38, mm10, rn6, etc.).  Also,
right click on the track and  color by first of pair strand to check if the first
read aligns in the opposite direction of the feature (note red is the positive
strand  and blue is negative).   This is usually the case for stranded Illumina
paired end sequencing runs so the `-s 2` option in `featureCounts` was used to
count reversely stranded reads.

To compare many samples, it's easier to load the bigWig files in GNomEx. The
STAR aligner normalizes coverage files by default, so the units are reads per
million mapped reads (RPM). In addition, separate coverage files are created for
unique reads (\*.unique.bw) and unique plus multi-mapping reads
(\*.multiple.bw).

[IGV] displays a normalized coverage plot and RPM values at each position.
If needed, multiply the RPM by the number of unique alignments from
STAR in the [MultiQC] file to get the total reads at that base (for example,
24.45863 RPM * 21873674 uniquely aligned reads/1000000 = 535 total reads).
Finally, select multiple tracks and right click to set the same minimun and
maximum values to display in the y-axis range by selecting Group Autoscale.

### 10. Differential Expression

See the `DESeq.html` report for further details on the differential expression analysis using
[DESeq2].

<br>

[problem]: https://sequencing.qcfail.com/articles/illumina-patterned-flow-cells-generate-duplicated-sequences/
[clumpify]: https://jgi.doe.gov/data-and-tools/bbtools/bb-tools-user-guide/clumpify-guide/
[DESeq2]: http://www.bioconductor.org/packages/release/bioc/html/DESeq2.html
[RSEM]: https://deweylab.github.io/RSEM/
[cutadapt]: http://cutadapt.readthedocs.io/en/stable/index.html
[STAR]: https://github.com/alexdobin/STAR
[Ensembl]: http://uswest.ensembl.org/info/data/ftp/index.html
[Pysano]: https://uofuhealth.utah.edu/huntsman/shared-resources/gba/bioinformatics/infrastructure/pysano.php
[CHPC]: https://www.chpc.utah.edu/
[shell]: http://tldp.org/LDP/abs/html/index.html
[IGV]: http://software.broadinstitute.org/software/igv/
[featureCounts]: http://bioinf.wehi.edu.au/featureCounts/
[FastQC]: http://www.bioinformatics.babraham.ac.uk/projects/fastqc/
[CollectRnaSeqMetrics]: https://broadinstitute.github.io/picard/command-line-overview.html#CollectRnaSeqMetrics
[samtools idxstats]: http://www.htslib.org/
[MultiQC]: http://multiqc.info/
